---
title: 'Reproducible Research: Peer Assessment 1'
author: "Wolfgang Ecker-Lala"
output:
  html_document:
    keep_md: yes
  word_document: default
---


## Loading and preprocessing the data
Data will be loaded from the datasource file named "activity.csv". This file has been downloaded and imported by
```{r,echo=TRUE}

filename <- "C:/MATH-UP/DataScience/Coursera DataScientist/Reproducible Research/Assignment1/activity.csv"

myData <- read.csv(filename,stringsAsFactor=FALSE)
```

In order to get for field "date" datatype POSIXct we make following transformation:
Data will be loaded from the datasource file named "activity.csv". This file has been downloaded and imported by
```{r,echo=TRUE}
library(lubridate)
myData$date<-ymd(myData$date)
```

## What is mean total number of steps taken per day?
The total number of steps taken per day is calculated by

```{r,echo=TRUE}
dailyActivity<-aggregate(steps~date,data=myData,FUN=sum)
```

and immediately we get an histogram by following command

```{r,echo=TRUE}
plot(dailyActivity$date,dailyActivity$steps,type="h",ylab="steps",xlab="Date",main="steps per day")
```

The calculated mean of daily sums of steps can be retrieved by

```{r,echo=TRUE}
stepMean<-mean(dailyActivity$steps)
```

So for the value of this mean we get `r format(stepMean,digits=10)`.

In same way the calculated median of sums of steps can be retrieved by

```{r,echo=TRUE}
stepMedian<-median(dailyActivity$steps)
```

So for the value of this median we get `r format(stepMedian,digits=10)`.

## What is the average daily activity pattern?

In order to be able to answer this question we have to do following transformation and calculation before:
```{r,echo=TRUE}
dailyActivityInterval<-aggregate(steps~interval,data=myData,FUN=mean)
allDaysMean<-mean(myData$steps,na.rm=TRUE)
```

So the daily activity pattern is like

```{r,echo=TRUE}
plot(dailyActivityInterval$interval,dailyActivityInterval$steps,type="l",xlab="interval",ylab="steps",main="average steps per interval",ylim=range(dailyActivityInterval$steps))
par(new=TRUE)
plot(c(min(dailyActivityInterval$interval),max(dailyActivityInterval$interval)),c(allDaysMean,allDaysMean),axes=FALSE,type="l",col="RED",xlab="",ylab="",main="",ylim=range(dailyActivityInterval$steps))
```

The red line there shows the average of steps across all days.

Now we like to get the interval on average across all the days with maximum number of steps.
For this we do following calculation.

```{r,echo=TRUE}
maxIntervalSteps<-max(dailyActivityInterval$steps)
maxIntervalBySteps<-dailyActivityInterval[dailyActivityInterval$steps==maxIntervalSteps,]
```

And the value of the interval with maximum number of steps is `r maxIntervalBySteps$interval`.

## Imputing missing values

The total number of missing values can be calculated by
```{r,echo=TRUE}
recordsNA<-length(myData[is.na(myData$steps) | is.na(myData$interval) | is.na(myData$date) ,1])
```
and the value is `r recordsNA`.

Missing values can be filled using following strategy:

Use the average value of steps of the corresponding interval over all observations.

A new dataset can be created by
```{r,echo=TRUE}
myNewData<-myData
missingSteps<-which(is.na(myData$steps)==TRUE)
for (i in 1:length(missingSteps)) {
  myNewData[missingSteps[i],]$steps<-dailyActivityInterval[dailyActivityInterval$interval==myNewData[missingSteps[i],]$interval,]$steps
}
```

In order to be able to plot again a histogram with these modified data we have to do following calculation before.
```{r,echo=TRUE}
dailyNewActivity<-aggregate(steps~date,data=myNewData,FUN=sum)
```

And create the histogram by
```{r,echo=TRUE}
plot(dailyNewActivity$date,dailyNewActivity$steps,type="h",ylab="steps",xlab="Date",main="steps per day")
```

Mean and median of this modified dataset can be calculated by
```{r,echo=TRUE}
stepNewMean<-mean(dailyNewActivity$steps)
stepNewMedian<-median(dailyNewActivity$steps)
```

And we get following values:

Mean: `r format(stepNewMean,digits=10)`

Median: `r format(stepNewMedian,digits=10)`

These values differ of course from mean and median of original data. Mean and median are equal now.

## Are there differences in activity patterns between weekdays and weekends?

We introduce a new field named "dayOfWeek" which can have the value *weekday* (monday, tuesday, wednesday,thursday,friday) or *weekend* (saturday, sunday).

```{r,echo=TRUE}
myNewData$dayOfWeek$dayOfWeek<-"weekday"
myNewData[weekdays(myNewData$date)=="Sonntag" | weekdays(myNewData$date)=="Samstag",]$dayOfWeek<-"weekend"
```

In order to be able to show the differences between activities of weekday and weekend we do following precalculation.

```{r,echo=TRUE}
weekdayData<-myNewData[myNewData$dayOfWeek=="weekday",]
weekendData<-myNewData[myNewData$dayOfWeek=="weekend",]
dailyWeekdayInterval<-aggregate(steps~interval,data=weekdayData,FUN=mean)
dailyWeekendInterval<-aggregate(steps~interval,data=weekendData,FUN=mean)
```

The following panel plot shows now the differences.

```{r,echo=TRUE}
par(mfrow=c(2,1))
plot(dailyWeekendInterval$interval,dailyWeekendInterval$steps,type="l",xlab="interval",ylab="number of steps",main="weekend",ylim=range(dailyWeekendInterval$steps))
plot(dailyWeekdayInterval$interval,dailyWeekdayInterval$steps,type="l",xlab="interval",ylab="number of steps",main="weekday",ylim=range(dailyWeekdayInterval$steps))
```
